---
title: "Predicting Flight Departure Delays from Scheduled Characteristics, Origin Weather, and the Knock-On Effect"
author: "Erel Edry, Idan Hazan, Nadav Zvulun, Reem Mizrahi"
subtitle: "Introduction to Data Science"
output:
  pdf_document:
    latex_engine: xelatex
    number_sections: false
    fig_caption: true
  html_document:
    df_print: paged
geometry: margin=0.3in
fontsize: 10pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.align = "center", fig.width = 6.0, fig.height = 3.8)
library(tidyverse)
set.seed(42)
```
<style>
/* tighten ### (h3) heading spacing by ~25% vs bootstrap defaults (20px/10px) */
h3 { margin-top: 15px; margin-bottom: 7.5px; line-height: 0.9; }
</style>
```{r data-load}
flights_raw <- read_csv("data/flight_with_weather.csv", show_col_types = FALSE)
tb_levels <- c("Morning (06:00-12:00)", "Afternoon (12:00-18:00)",
               "Evening (18:00-24:00)", "Night (00:00-06:00)")
major_airlines <- c("United Air Lines Inc.", "American Airlines Inc.",
                    "US Airways Inc.", "Delta Air Lines Inc.",
                    "Alaska Airlines Inc.", "Hawaiian Airlines Inc.")
flights <- flights_raw %>%
  mutate(
    is_delayed     = if_else(DEPARTURE_DELAY > 15, 1, 0),
    hour           = as.integer(SCHEDULED_DEPARTURE) %/% 100,
    Time_Block     = case_when(
      hour >= 6  & hour < 12 ~ "Morning (06:00-12:00)",
      hour >= 12 & hour < 18 ~ "Afternoon (12:00-18:00)",
      hour >= 18 & hour < 24 ~ "Evening (18:00-24:00)",
      TRUE                   ~ "Night (00:00-06:00)"),
    MONTH          = factor(MONTH),
    DAY_OF_WEEK    = factor(DAY_OF_WEEK),
    Time_Block     = factor(Time_Block, levels = tb_levels),
    Carrier_Type   = if_else(str_trim(AIRLINE) %in% major_airlines, "Major", "Low-Cost"),
    Carrier_Type   = factor(Carrier_Type),
    Bad_Weather    = if_else(!is.na(ORIGIN_PRESENT_WEATHER_CODES), 1, 0),
    prev_arr_delay = if_else(is.na(PREV_LEG_ARR_DELAY), 0, PREV_LEG_ARR_DELAY),
    has_prev_leg   = if_else(is.na(PREV_LEG_ARR_DELAY), 0L, 1L),
    Prev_Leg       = factor(case_when(
      is.na(PREV_LEG_ARR_DELAY) ~ "First leg of day",
      PREV_LEG_ARR_DELAY > 15   ~ "Inbound late (>15)",
      TRUE                      ~ "Inbound on-time"),
      levels = c("First leg of day", "Inbound on-time", "Inbound late (>15)"))
  ) %>%
  drop_na(DEPARTURE_DELAY, DISTANCE, DAY_OF_WEEK, Carrier_Type,
          ORIGIN_WIND_SPEED_KNOTS, Bad_Weather, prev_arr_delay, has_prev_leg,
          TMAX, TMIN, PRCP)
set.seed(42)
flights    <- flights %>% slice_sample(n = 200000)   # uniform random sample of 200,000 complete-case flights
delay_rate <- mean(flights$is_delayed)
```
```{r modelling}
set.seed(42)
n         <- nrow(flights)
train_idx <- sample(seq_len(n), size = floor(0.8 * n))
train     <- flights %>% slice(train_idx)
test      <- flights %>% slice(-train_idx)

# Base OLS — scheduled features only, on the SAME 200K split (clean benchmark;
# computed here so r2_base is available to the inline reference in Section 1)
model_base <- lm(DEPARTURE_DELAY ~ DISTANCE + Time_Block + DAY_OF_WEEK + Carrier_Type, data = train)
pred_base  <- predict(model_base, newdata = test)
r2_base    <- 1 - sum((test$DEPARTURE_DELAY - pred_base)^2, na.rm = TRUE) /
                  sum((test$DEPARTURE_DELAY - mean(test$DEPARTURE_DELAY, na.rm = TRUE))^2, na.rm = TRUE)

potential_vars <- c("DISTANCE", "Time_Block", "DAY_OF_WEEK", "Carrier_Type",
                    "Bad_Weather", "ORIGIN_WIND_SPEED_KNOTS", "prev_arr_delay",
                    "TMAX", "TMIN", "PRCP")
valid_vars <- potential_vars[sapply(potential_vars, function(v) {
  length(unique(train[[v]][!is.na(train[[v]])])) > 1
})]
glm_formula <- as.formula(paste("is_delayed ~", paste(valid_vars, collapse = " + ")))
model_glm   <- glm(glm_formula, family = binomial, data = train)
test        <- test %>%
  mutate(pred_prob_glm = predict(model_glm, newdata = test, type = "response"))
roc <- test %>% arrange(desc(pred_prob_glm)) %>%
  mutate(TPR = cumsum(is_delayed == 1) / sum(is_delayed == 1),
         FPR = cumsum(is_delayed == 0) / sum(is_delayed == 0))
auc      <- sum(diff(roc$FPR) * (head(roc$TPR, -1) + tail(roc$TPR, -1)) / 2)
test     <- test %>% mutate(pred_class_glm = if_else(pred_prob_glm > delay_rate, 1, 0))
accuracy_glm <- mean(test$pred_class_glm == test$is_delayed)
tp    <- sum(test$is_delayed == 1 & test$pred_class_glm == 1, na.rm = TRUE)
fp    <- sum(test$is_delayed == 0 & test$pred_class_glm == 1, na.rm = TRUE)
fn    <- sum(test$is_delayed == 1 & test$pred_class_glm == 0, na.rm = TRUE)
p_glm   <- if_else((tp + fp) == 0, 0, tp / (tp + fp))
r_glm   <- if_else((tp + fn) == 0, 0, tp / (tp + fn))
mse_glm <- mean((test$is_delayed - test$pred_prob_glm)^2, na.rm = TRUE)
base_risk <- mean(flights$is_delayed[flights$has_prev_leg == 0])
```

### Section 1 — Introduction

Flight delays cost the US economy upward of $28 billion annually (Ball et al., *Total Delay Impact Study*, NEXTOR 2010) through crew overtime, fuel burn, gate occupancy, and the downstream burden of missed connections. Crucially, delays do not stay local: a single disrupted flight propagates across a hub-and-spoke network, cascading through later legs flown by the same aircraft. This propagation is the **knock-on (domino) effect** — because each departure depends on the timely arrival of the inbound aircraft, a late inbound compresses the turnaround window, and any shortfall converts directly into an outbound delay that compounds across the day — a propagation mechanism documented for passenger airline networks by AhmadBeygi, Cohn, Guan & Belobaba (*Analysis of the potential for delay propagation in passenger airline networks*, J. Air Transport Management, 2008).

Prior models, including our own, relied only on scheduled features (carrier, distance, time-of-day): on our 200K hold-out a scheduled-only OLS explains just **`r round(r2_base*100,1)`%** of delay variance (held-out R² = **`r round(r2_base,3)`**), with a logistic AUC of 0.631. That ceiling was a _data gap_, not a modelling failure: without inbound arrival performance and origin weather, the two dominant physical drivers were absent. This report adds both. 

### Section 2 — Data Overview

**Sampling & cleaning.** From the 5.8M-flight 2015 DOT dataset we join NOAA daily weather by origin airport and date, drop cancellations (missing `DEPARTURE_DELAY` , structurally distinct from delays), and keep complete cases across all model features. From this cleaned pool we draw a uniform random sample of **`r format(nrow(flights), big.mark=",")`** flights for analysis, **`r round(delay_rate*100,1)`%** of them delayed (DEPARTURE_DELAY > 15 min, the DOT threshold). This imbalance motivates the calibrated classification threshold in Section 3. 

**Features.** Five groups: _Temporal_ (`MONTH`, `DAY_OF_WEEK`, `Time_Block`); _Carrier & Distance_ (Major/Low-Cost split, route miles); _Weather_ (`Bad_Weather`, `ORIGIN_WIND_SPEED_KNOTS`, `TMAX`, `TMIN`, `PRCP`); _Knock-on_ (`prev_arr_delay` = inbound arrival delay in minutes, set to 0 for first legs, with a `has_prev_leg` flag so the model can separate “no inbound constraint” from “on-time inbound”); and _targets_ `DEPARTURE_DELAY` (continuous) and `is_delayed`. 

### Section 3 — Methods and Results

**Architecture.** On an 80/20 train/test split (seed 42) we fit OLS on `DEPARTURE_DELAY` (delay magnitude) and logistic regression on `is_delayed` (delay occurrence) over an identical predictor set, so coefficient signs can be cross-checked across both targets. 

**Feature importance (Figure 1).** Re-fitting OLS on z-standardised inputs, `prev_arr_delay` dominates — its scaled coefficient is several times any other predictor, confirming inbound performance as the primary driver of outbound risk. Evening time-blocks form a second tier (cumulative knock-on burden); the weather cluster sits mid-tier; distance and carrier type are weakest, consistent with delay being a ground-side phenomenon. 

```{r fig-importance, fig.width=4.0, fig.height=2.1}
train_scaled <- train %>%
  mutate(across(any_of(c("DISTANCE", "ORIGIN_WIND_SPEED_KNOTS",
                          "prev_arr_delay", "TMAX", "TMIN", "PRCP")),
                ~ as.numeric(scale(.))))
model_lm_scaled <- lm(as.formula(paste("DEPARTURE_DELAY ~",
                      paste(valid_vars, collapse = " + "))), data = train_scaled)
var_imp <- abs(coef(model_lm_scaled)[-1])
imp_df  <- tibble(variable = names(var_imp), importance = var_imp) %>%
           arrange(desc(importance)) %>% slice_max(importance, n = 7)
ggplot(imp_df, aes(x = reorder(variable, importance), y = importance)) +
  geom_col(fill = "steelblue") + coord_flip() +
  labs(title = "Figure 1. Feature Importance",
       subtitle = "Top 7 predictors",
       x = "Predictor", y = "Absolute Scaled Coefficient") +
  theme_minimal(base_size = 10) +
  theme(plot.title    = element_text(face = "bold", hjust = 0.5, size = 10),
        plot.subtitle = element_text(hjust = 0.5, color = "gray40", size = 9))
```

**Carrier, weather & domino dynamics.** The empirical delay rate is **`r round(mean(flights$is_delayed[flights$Carrier_Type=="Low-Cost"])*100,1)`%** for Low-Cost vs **`r round(mean(flights$is_delayed[flights$Carrier_Type=="Major"])*100,1)`%** for Major carriers (negative `Carrier_TypeMajor` log-odds — larger turnaround buffers). Wind speed carries a positive coefficient (crosswind pushback/runway delays), and Low-Cost carriers absorb a larger weather penalty given their tighter schedules. The domino curve (Figure 3) is the key result: first-leg risk is **`r round(base_risk*100,1)`%**, but rises steeply and non-linearly with inherited inbound lateness — ~30 min roughly doubles baseline risk, accelerating toward near-certain delay beyond that. 

**Performance.** With only `r round(delay_rate*100,1)`% delayed, a 0.5 cutoff would label almost everything on-time; the threshold is instead fixed at the empirical delay rate (Table 1). The AUC gain over the 0.631 baseline confirms weather and knock-on features add independent signal. 

Table 1. Held-out performance metrics (20% test set). 

```{r metrics-table}
tibble(Model     = "Logistic Regression",
       Accuracy  = round(accuracy_glm, 4), AUC = round(auc, 4),
       Precision = round(p_glm, 4), Recall = round(r_glm, 4),
       MSE       = round(mse_glm, 4)) %>%
  knitr::kable()
```

```{r fig-roc-domino, fig.width=5.0, fig.height=1.8, out.width='80%'}
library(patchwork)
p_roc <- ggplot(roc, aes(x = FPR, y = TPR)) +
  geom_line(color = "#2C7FB8", linewidth = 0.8) +
  geom_abline(linetype = "dashed", color = "grey50") +
  labs(title = paste0("Figure 2. ROC Curve AUC = ", round(auc, 3), ""),
       x = "False Positive Rate", y = "True Positive Rate") +
  theme_minimal(base_size = 7) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 8))

p_domino <- flights %>% filter(has_prev_leg == 1) %>%
  ggplot(aes(x = prev_arr_delay, y = is_delayed)) +
  geom_hline(yintercept = base_risk, linetype = "dashed", color = "#2CA25F", linewidth = 1.0) +
  annotate("text", x = 90, y = base_risk + 0.06,
           label = paste0("Baseline: ", round(base_risk * 100, 1), "%"),
           color = "#2CA25F", fontface = "bold", size = 2.6) +
  stat_smooth(method = "glm", method.args = list(family = "binomial"),
              color = "#E34A33", linewidth = 1.0, se = TRUE, fill = "pink") +
  coord_cartesian(xlim = c(-15, 150)) +
  scale_y_continuous(labels = scales::percent_format(), breaks = seq(0, 1, 0.2)) +
  scale_x_continuous(breaks = seq(-15, 150, 30)) +
  labs(title = "Figure 3. The Domino Effect", subtitle = "Inbound delay vs. departure delay risk",
       x = "Inbound Flight Delay (min)", y = "P(Departure Delay > 15 min)") +
  theme_minimal(base_size = 7) +
  theme(plot.title       = element_text(face = "bold", hjust = 0.5, size = 8),
        plot.subtitle    = element_text(hjust = 0.5, color = "gray40", size = 7),
        panel.grid.minor = element_blank())

p_roc + p_domino
```

## Model Development: Advanced Modeling & Goodness-of-Fit

We now extend the baseline models above to address two OLS limitations — the right-skewed, near-zero `prev_arr_delay` and non-constant error variance — by (1) log-transforming `prev_arr_delay`, (2) re-fitting OLS with interaction terms, (3) checking residual structure, and (4) fitting a GLM with a log link. This model-development and fit-checking work is the core of Section 3’s “how well does the model fit the data” requirement. 

**Log-scaling** **`prev_arr_delay`.** `prev_arr_delay` mixes structural zeros (first legs) with a long positive tail; a log transform compresses that tail and linearises the inbound→outbound relationship (diminishing marginal impact of each extra inherited minute). Since values can be ≤0, we shift by $c = |\min(\text{prev\_arr\_delay})| + 2$ so the argument exceeds 1, then take logs. 

**Extended OLS with interactions.** We replace raw `prev_arr_delay` with its log and add three two-way interactions — `log_prev × Carrier_Type` (does knock-on transmit differently by carrier?), `log_prev × Bad_Weather` (does weather compound inbound lateness?), and `Carrier_Type × Bad_Weather` (carrier weather vulnerability). 

```{r ext-ols}
# 0. יצירת המשתנה הלוגריתמי כפי שהוגדר בטקסט
c_shift <- abs(min(train$prev_arr_delay, na.rm = TRUE)) + 2
train <- train %>% mutate(log_prev_arr_delay = log(prev_arr_delay + c_shift))
test <- test %>% mutate(log_prev_arr_delay = log(prev_arr_delay + c_shift))

# 1. חישוב הממוצע של משתנה הלוג (על קבוצת האימון) וביצוע המרכוז
mean_log_delay <- mean(train$log_prev_arr_delay, na.rm = TRUE)
train <- train %>% mutate(log_prev_arr_delay_centered = log_prev_arr_delay - mean_log_delay)
test <- test %>% mutate(log_prev_arr_delay_centered = log_prev_arr_delay - mean_log_delay)

# 2. הרצת מודל OLS עם המשתנה הממורכז
model_ols_ext <- lm(
  DEPARTURE_DELAY ~ log_prev_arr_delay_centered + Carrier_Type + Bad_Weather +
    ORIGIN_WIND_SPEED_KNOTS + TMAX + TMIN + PRCP + Time_Block + DAY_OF_WEEK + DISTANCE +
    log_prev_arr_delay_centered:Carrier_Type + log_prev_arr_delay_centered:Bad_Weather +
    Carrier_Type:Bad_Weather,
  data = train
)

# חישוב R² על מדגם המבחן (held-out)
pred_ext <- predict(model_ols_ext, newdata = test)
r2_ext   <- 1 - sum((test$DEPARTURE_DELAY - pred_ext)^2, na.rm = TRUE) /
                sum((test$DEPARTURE_DELAY - mean(test$DEPARTURE_DELAY, na.rm = TRUE))^2, na.rm = TRUE)

# 3. חילוץ מקדמים אמיתיים מהמודל לטבלה דינמית
ext_terms <- c("log_prev_arr_delay_centered", "Carrier_TypeMajor", "Bad_Weather",
               "log_prev_arr_delay_centered:Carrier_TypeMajor",
               "log_prev_arr_delay_centered:Bad_Weather", "Carrier_TypeMajor:Bad_Weather")

co_ext <- summary(model_ols_ext)$coefficients
co_ext_filtered <- co_ext[rownames(co_ext) %in% ext_terms, , drop = FALSE]

tibble(
  Term = rownames(co_ext_filtered),
  Estimate = round(co_ext_filtered[, 1], 3),
  SE = round(co_ext_filtered[, 2], 3),
  `t-value` = round(co_ext_filtered[, 3], 2),
  `p-value` = formatC(co_ext_filtered[, 4], format = "e", digits = 2)
) %>%
  knitr::kable(caption = "Table 2. Extended OLS — key coefficients.")
```

On the **same** 200K hold-out, a Base OLS using only scheduled features (distance, time-of-day, day, carrier) reaches held-out R² = **`r round(r2_base,3)`**; adding the knock-on channel and the interaction structure lifts it to **`r round(r2_ext,3)`** — a like-for-like gain on identical data, not a cross-pipeline figure (full Base / Extended / GLM comparison in Table 4). _Note:_ To ensure the main effects remain interpretable despite the interaction terms, `log_prev_arr_delay` was mean-centered. Consequently, the coefficient for `Bad_Weather` now accurately represents the isolated delay penalty of adverse weather for a flight experiencing an average inbound delay, resolving the negative-coefficient artifact of an uncentered interaction model.

**Residual diagnostics.** We check homoscedasticity and normality on the training residuals of the extended model (Figures 5–6).

```{r s5-diagnostics, fig.width=5.0, fig.height=1.6, out.width='80%'}
library(patchwork)
diag_df <- tibble(fitted = fitted(model_ols_ext), residuals = residuals(model_ols_ext))
diag_df_sampled <- diag_df %>% sample_n(min(10000, nrow(diag_df)))
p1 <- ggplot(diag_df_sampled, aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.15, size = 0.3, color = "#2C7FB8") +
  geom_hline(yintercept = 0, color = "firebrick", linewidth = 0.4, linetype = "dashed") +
  geom_smooth(method = "loess", se = FALSE, color = "#E34A33", linewidth = 0.5, span = 0.4) +
  coord_cartesian(ylim = c(-60, 120)) +
  labs(title = "Fig 5. Residuals vs Fitted", subtitle = "Homoscedasticity check",
       x = "Fitted (min)", y = "Residuals (min)") +
  theme_minimal(base_size = 6)
p2 <- ggplot(diag_df, aes(x = residuals)) +
  geom_histogram(aes(y = after_stat(density)), bins = 80, fill = "steelblue",
                 color = "white", alpha = 0.85) +
  stat_function(fun = dnorm, args = list(mean = mean(diag_df$residuals),
                sd = sd(diag_df$residuals)), color = "firebrick", linewidth = 0.6) +
  coord_cartesian(xlim = c(-80, 160)) +
  labs(title = "Fig 6. Histogram of Residuals", subtitle = "Normality check",
       x = "Residual (min)", y = "Density") +
  theme_minimal(base_size = 6)
p1 + p2
```

Residual spread fans out with fitted value (**heteroscedasticity**), a structural feature of delay data (bounded below near −30 min, unbounded above), while the LOESS line stays near zero — so OLS estimates remain unbiased but standard errors are unreliable (robust SE or a variance-stabilising GLM are the fix). Residuals are strongly right-skewed, so normal prediction intervals are too narrow in the right tail; point estimates stay valid via the CLT. 

**GLM with log link.** To handle skew in a principled way we fit a Gamma GLM with a **log link**: $E[\text{delay} | X] = \exp(X\beta)$, so predictors act multiplicatively and fitted means stay non-negative. The response is shifted by $c_y = |\min(\text{DEPARTURE\_DELAY})| + 1$ to keep the link argument positive. 

```{r s5-glm-dynamic}
# 1. הזזת משתנה המטרה והרצת מודל GLM דינמי
c_y <- abs(min(train$DEPARTURE_DELAY, na.rm = TRUE)) + 1
train_glm <- train %>% mutate(DEP_DELAY_SHIFT = DEPARTURE_DELAY + c_y)

model_glm_log <- glm(
  DEP_DELAY_SHIFT ~ log_prev_arr_delay + Carrier_Type + Bad_Weather +
    ORIGIN_WIND_SPEED_KNOTS + TMAX + TMIN + PRCP + Time_Block + DAY_OF_WEEK + DISTANCE,
  family = Gamma(link = "log"),
  data = train_glm,
  control = glm.control(maxit = 100)
)

# חישוב RMSE על מדגם המבחן (DEP_DELAY_SHIFT הוא ה-response, לא predictor — לכן test הרגיל מספיק)
pred_glm_orig <- predict(model_glm_log, newdata = test, type = "response") - c_y
rmse_glm_log  <- sqrt(mean((test$DEPARTURE_DELAY - pred_glm_orig)^2, na.rm = TRUE))
rmse_ols_ext  <- sqrt(mean((test$DEPARTURE_DELAY - pred_ext)^2, na.rm = TRUE))

# 2. חילוץ דינמי של מקדמי ה-GLM לטבלה
glm_terms <- c("(Intercept)", "log_prev_arr_delay", "Carrier_TypeMajor", "Bad_Weather", "ORIGIN_WIND_SPEED_KNOTS", "PRCP")
co_glm <- summary(model_glm_log)$coefficients
co_glm_filtered <- co_glm[rownames(co_glm) %in% glm_terms, , drop = FALSE]

tibble(
  Term = rownames(co_glm_filtered),
  Estimate = round(co_glm_filtered[, 1], 4),
  SE = round(co_glm_filtered[, 2], 4),
  `t-value` = round(co_glm_filtered[, 3], 2),
  `p-value` = formatC(co_glm_filtered[, 4], format = "e", digits = 2)
) %>%
  knitr::kable(caption = "Table 3. GLM (Gamma, log link) — key coefficients.")
```

Each coefficient is a _multiplicative_ effect: $\exp(\hat{\beta})$ is the factor by which expected delay changes per unit (held-out RMSE **`r round(rmse_glm_log,2)`** vs **`r round(rmse_ols_ext,2)`** min for the extended OLS). We prefer the GLM on empirical grounds: on the identical 200K hold-out it wins on all three metrics (R², RMSE, MAE; Table 4) and guarantees non-negative fitted means. Its theoretical motivation — Gamma's variance $\propto$ mean$^2$ matching the heteroscedasticity in the diagnostics (Figures 5–6) — is only partial here, since the additive shift $c_y$ compresses the mean's dynamic range and thus attenuates that variance structure.

All three regression models are evaluated on the **same** 200K hold-out with identical metrics (Table 4), so the improvement is a clean apples-to-apples comparison rather than a figure carried over from an earlier pipeline.

```{r benchmark-table}
metrics <- function(actual, pred) {
  ss_tot <- sum((actual - mean(actual, na.rm = TRUE))^2, na.rm = TRUE)
  tibble(R2   = 1 - sum((actual - pred)^2, na.rm = TRUE) / ss_tot,
         RMSE = sqrt(mean((actual - pred)^2, na.rm = TRUE)),
         MAE  = mean(abs(actual - pred), na.rm = TRUE))
}
y <- test$DEPARTURE_DELAY
bind_rows(
  tibble(Model = "Base OLS (scheduled only)")         %>% bind_cols(metrics(y, pred_base)),
  tibble(Model = "Extended OLS (log + interactions)") %>% bind_cols(metrics(y, pred_ext)),
  tibble(Model = "GLM (Gamma, log link)")             %>% bind_cols(metrics(y, pred_glm_orig))
) %>%
  mutate(across(c(R2, RMSE, MAE), ~ round(., 3))) %>%
  knitr::kable(caption = "Table 4. Same-data model comparison — 200K hold-out (20% test): held-out R², RMSE (min) and MAE (min).")
```

### Section 4 — Limitations and Future Work

Structural gaps remain: airport ground congestion and ATC ground-stops are absent from DOT records; the 2015-only scope misses multi-year drift; and our continuous weather features (TMAX/TMIN/PRCP) are **daily** NOAA summaries, while `Bad_Weather` is built from origin present-weather codes (`ORIGIN_PRESENT_WEATHER_CODES`). None of these is matched to the exact departure hour, so sub-daily timing of conditions is not captured. **Future work:** ensemble non-linear models (Random Forest, XGBoost) to capture weather–network interactions; multi-year data; hourly METAR weather joined to departure time; and FAA ASPM congestion features.