---
title: "Predicting Flight Departure Delays from Scheduled Characteristics"
subtitle: "Introduction to Data Science"
author: "Erel Edry, Idan Hazan, Nadav Zvulun, Reem Mizrahi"
output:
  pdf_document:
    number_sections: false
    fig_caption: true
geometry: margin=0.9in
fontsize: 11pt
---

```{r setup, include=FALSE}
# Course rule: tidyverse (and data.table) only for data work; models use base R.
# echo=FALSE keeps the report clean but fully reproducible.
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.align = "center", fig.width = 4.0, fig.height = 2.5)
library(tidyverse)
set.seed(42)   # makes the 100k sample and all results reproducible
```

**Code repository:** `(see `README.md` to reproduce).\
**Data:** US DOT 2015 domestic flights staff-provided, not in the report.

```{r load-clean, cache=TRUE}
# Read only the columns we need from the large (~5.8M row) flights file.
flights_raw <- read_csv(
  "data/flights.csv",
  col_select = c(MONTH, DAY_OF_WEEK, AIRLINE,
                 SCHEDULED_DEPARTURE, DEPARTURE_DELAY, DISTANCE),
  show_col_types = FALSE
)

# Rename the airlines name column so it doesn't clash with the flights code column.
airlines <- read_csv("data/airlines.csv", show_col_types = FALSE) %>%
  rename(Airline_Name = AIRLINE)

# 100k sample, joined to names, cancelled flights removed (missing delay = cancelled).
flights <- flights_raw %>%
  sample_n(100000) %>%
  left_join(airlines, by = c("AIRLINE" = "IATA_CODE")) %>%
  filter(!is.na(DEPARTURE_DELAY))

# Time_Block levels carry the explicit hour ranges (used on axes + coefficient table).
tb_levels <- c("Morning (06:00-12:00)", "Afternoon (12:00-18:00)",
               "Evening (18:00-24:00)", "Night (00:00-06:00)")

major_airlines <- c("UA", "AA", "US", "DL", "AS", "HA")   # legacy carriers
flights <- flights %>%
  mutate(
    is_delayed = if_else(DEPARTURE_DELAY > 15, 1, 0),
    # SCHEDULED_DEPARTURE is text like "0545": convert, then take the hour.
    hour = as.integer(SCHEDULED_DEPARTURE) %/% 100,
    Time_Block = case_when(
      hour >= 6  & hour < 12 ~ "Morning (06:00-12:00)",
      hour >= 12 & hour < 18 ~ "Afternoon (12:00-18:00)",
      hour >= 18 & hour < 24 ~ "Evening (18:00-24:00)",
      TRUE                   ~ "Night (00:00-06:00)"),
    MONTH        = factor(MONTH),
    DAY_OF_WEEK  = factor(DAY_OF_WEEK),
    Time_Block   = factor(Time_Block, levels = tb_levels),
    Carrier_Type = if_else(AIRLINE %in% major_airlines, "Major", "Low-Cost")
  )

delay_rate <- mean(flights$is_delayed)
cr <- flights %>% group_by(Carrier_Type) %>%
  summarise(r = mean(is_delayed), .groups = "drop")
rate_low   <- cr$r[cr$Carrier_Type == "Low-Cost"]
rate_major <- cr$r[cr$Carrier_Type == "Major"]
```

# Section 1 - Introduction

Flight delays cost airlines billions of dollars a year and frustrate millions of passengers.
They push airport schedules off track, force extra spending on crew, fuel, and gate time, and
make air travel hard to plan. Our research question is: how do a flight's scheduled features
distance, time of day, day of week, and carrier affect its departure delay, and can we predict
whether a flight will be delayed by more than 15 minutes? We use 15 minutes because that is the
cutoff the US DOT uses to label a flight as delayed.

This is hard for two reasons. First, many causes act at the same time weather, busy travel
seasons, crowded peak hours, so a single feature on its own can mislead. Second, delays cascade:
the same aircraft and crew fly several legs in a row, so a late first flight tends to make the next
flights late too. This knock on effect means a small morning delay can grow into large delays by
the evening.

Prior work shows that timing and carrier carry real signal, but that scheduled features
alone explain only part of delays, because weather and the wider flight network matter a lot.
Building on this, we test two hypotheses. **H1:** low cost carriers are delayed more than major
legacy carriers, since they fly tighter schedules with less spare time. **H2:** delays grow over
the course of the day, following the knock on effect. We use two models a linear regression for
how long a delay is, and a logistic regression for whether a flight is delayed and we show how
two careless choices removing too many outliers, using a 0.5 cutoff can break the analysis, and
how we fixed them.

# Section 2 - Data overview

Each row is one US domestic flight from the 2015 DOT dataset: **5,819,080 flights** and 31 columns,
plus lookup tables for airline names and airports. We draw a random sample of **100,000 flights**
to keep the analysis fast while staying representative; a random sample keeps the same mix of
carriers, times, and delays as the full data. Flights with a missing DEPARTURE_DELAY were
**cancelled** they never took off, so we drop them these are not random gaps but a different
kind of event. After cleaning, **`r format(nrow(flights), big.mark=",")`** flights remain, and
**`r round(delay_rate*100, 1)`%** are delayed by more than 15 minutes about 1 in 6. We use three
feature groups plus the target (Table 1).

```{r tbl-features}
tibble(
  `Feature group` = c("Temporal", "Carrier", "Distance", "Target"),
  `Variables`     = c("MONTH, DAY_OF_WEEK, Time_Block (Morning 06-12, Afternoon 12-18, Evening 18-24, Night 00-06)",
                      "Carrier_Type (Low-Cost vs Major)",
                      "DISTANCE (miles)",
                      "DEPARTURE_DELAY (min); is_delayed (1 if > 15 min)"),
  `Type`          = c("categorical","categorical","numeric","numeric / binary")
) %>%
  knitr::kable(caption = "Table 1. Features used in the analysis.")
```

splits the scheduled hour into Morning (06:00-12:00), Afternoon (12:00-18:00),
Evening (18:00-24:00), and Night (00:00-06:00). Carrier_Type labels the six legacy carriers
(UA, AA, US, DL, AS, HA) as **Major** and all others as **Low Cost**. We do not use the airports
table.

# Section 3 - Methods and results

We run two models with the same predictors, one for each part of the question: a linear regression
(OLS) for *how long* the delay is, and a logistic regression for *whether* a flight is delayed.
Running both lets us check that the story holds under two different methods.

**Hypothesis 1 (carrier).** The empirical delay rate is
**`r round(rate_low*100, 1)`%** for Low Cost carriers versus **`r round(rate_major*100, 1)`%** for
Major carriers Figure 1. This gap is direct, real world proof behind the negative Major
coefficient in the logistic model: belonging to a major carrier lowers the chance of a delay.

```{r fig-prop, fig.width=4.0, fig.height=2.5, out.width="60%", fig.cap="Figure 1. Share of flights delayed (>15 min) by carrier type. Low-Cost is clearly higher, supporting H1."}
flights %>%
  group_by(Carrier_Type) %>%
  summarise(delay_rate = mean(is_delayed), .groups = "drop") %>%
  ggplot(aes(x = Carrier_Type, y = delay_rate, fill = Carrier_Type)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_text(aes(label = paste0(round(delay_rate * 100, 1), "%")),
            vjust = -0.4, size = 2.6) +
  scale_y_continuous(labels = function(x) paste0(x * 100, "%"),
                     expand = expansion(mult = c(0, 0.15))) +
  labs(title = "H1: Share of Flights Delayed (>15 min)",
       x = "Carrier Type", y = "Delayed flights") +
  theme_minimal(base_size = 8)
```

**Hypothesis 2 (time of day).** Figure 2 shows departure delay across the four time blocks. The
spread and upper whisker grow from morning through evening and drop again at night, matching the
knock on build up we expected.

```{r fig-time, fig.width=4.0, fig.height=2.5, out.width="66%", fig.cap="Figure 2. Departure delay by time block. Delays build from morning to evening and reset at night, supporting H2. Y-axis zoomed to +/-40 min."}
ggplot(flights, aes(x = Time_Block, y = DEPARTURE_DELAY, fill = Time_Block)) +
  geom_boxplot(outlier.size = 0.2, linewidth = 0.3, show.legend = FALSE) +
  coord_cartesian(ylim = c(-20, 40)) +
  scale_x_discrete(labels = function(x) str_wrap(x, 10)) +
  labs(title = "H2: Delay by Time of Day", x = NULL, y = "Delay (min)") +
  theme_minimal(base_size = 8) +
  theme(axis.text.x = element_text(size = 6))
```

```{r models}
# 80/20 train/test split so every reported score is on flights the model never saw.
set.seed(42)
n         <- nrow(flights)
train_idx <- sample(seq_len(n), size = floor(0.8 * n))
train     <- flights %>% slice(train_idx)
test      <- flights %>% slice(-train_idx)

# Linear model: remove extreme outliers with the IQR rule (from the TRAINING data
# only) -- ONLY here, since OLS is sensitive to them. Doing this on all the data
# would delete most delayed flights (delay rate drops to ~5.6%) and break the
# logistic model, so Model 2 uses the full data.
q <- quantile(train$DEPARTURE_DELAY, c(0.25, 0.75)); iqr <- q[2] - q[1]
lo <- q[1] - 1.5*iqr; hi <- q[2] + 1.5*iqr
train_lm <- train %>% filter(DEPARTURE_DELAY >= lo, DEPARTURE_DELAY <= hi)
model_lm <- lm(DEPARTURE_DELAY ~ DISTANCE + Time_Block + DAY_OF_WEEK + Carrier_Type,
               data = train_lm)
# Held-out R^2: score on the comparable (same IQR window) part of the test set.
test_lm <- test %>% filter(DEPARTURE_DELAY >= lo, DEPARTURE_DELAY <= hi) %>%
  mutate(pred = predict(model_lm, newdata = .))
r2_test <- 1 - sum((test_lm$DEPARTURE_DELAY - test_lm$pred)^2) /
               sum((test_lm$DEPARTURE_DELAY - mean(test_lm$DEPARTURE_DELAY))^2)

# Logistic model fitted on the training data, scored on the held-out test set.
model_glm <- glm(is_delayed ~ DISTANCE + Time_Block + DAY_OF_WEEK + Carrier_Type,
                 family = binomial, data = train)
test      <- test %>% mutate(pred_prob = predict(model_glm, newdata = test, type = "response"))
or_evening <- unname(exp(coef(model_glm)["Time_BlockEvening (18:00-24:00)"]))
or_major   <- unname(exp(coef(model_glm)["Carrier_TypeMajor"]))

# ROC + AUC by hand on the test set (tidyverse only).
roc <- test %>% arrange(desc(pred_prob)) %>%
  mutate(TPR = cumsum(is_delayed == 1)/sum(is_delayed == 1),
         FPR = cumsum(is_delayed == 0)/sum(is_delayed == 0))
auc <- sum(diff(roc$FPR) * (head(roc$TPR,-1) + tail(roc$TPR,-1))/2)

# Use the delay rate (~0.18) as the cutoff, not 0.5 (which calls everything on-time).
test     <- test %>% mutate(pred_class = if_else(pred_prob > delay_rate, 1, 0))
accuracy <- mean(test$pred_class == test$is_delayed)
```

**Coefficients.** Both models agree in sign and significance Table 2. Distance barely matters
about 1 minute per 1,000 miles, which fits the idea that departure delays build on the ground,
not in the air. Compared to morning, afternoon and evening flights have about +2 minutes of delay
while night flights are about 1 minute earlier (H2). The Major coefficient is negative, so major
carriers are delayed less than low cost carriers (H1). On the odds scale, evening flights have
about **`r round(or_evening, 2)`** times the odds of a delay versus morning, and major carriers
have about **`r round((1 - or_major) * 100)`%** lower odds than low cost carriers.

```{r tbl-coef}
k  <- c("DISTANCE", "Time_BlockAfternoon (12:00-18:00)",
        "Time_BlockEvening (18:00-24:00)", "Time_BlockNight (00:00-06:00)",
        "Carrier_TypeMajor")
lb <- c("DISTANCE (per mile)", "Afternoon (12:00-18:00)", "Evening (18:00-24:00)",
        "Night (00:00-06:00)", "Carrier_Type: Major")
tibble(Predictor = lb,
       `Linear (min)`        = round(coef(model_lm)[k], 4),
       `Logistic (log-odds)` = round(coef(model_glm)[k], 4),
       `Odds ratio`          = round(exp(coef(model_glm)[k]), 3)) %>%
  knitr::kable(row.names = FALSE,
    caption = "Table 2. Key coefficients (all p < 0.001). Baseline = Morning (06:00-12:00), Low-Cost.")
```

**Model fit.** We fit both models on 80% of the flights and report every score on the held-out 20%,
so the numbers reflect how the models do on flights they never saw. The linear model's held-out
$R^2$ is only **`r round(r2_test, 3)`**, so scheduled features explain about
`r round(r2_test*100, 1)`% of the variation in delay. We treat this as a real finding: the biggest
causes of delay weather, congestion, and the knock on effect are not in our data. For the
classifier, the held-out **AUC is `r round(auc, 3)`** (Figure 3), which is modest but clearly above
the 0.5 line for random guessing, so the features carry real signal. We are careful about accuracy:
with only about `r round(delay_rate*100)`% of flights delayed, a 0.5 cutoff would call almost every
flight on time and report a misleading accuracy near the base rate. Instead we set the cutoff at the
delay rate (`r round(delay_rate, 2)`), giving an honest accuracy of **`r round(accuracy, 3)`** from a
model that actually flags delayed flights.


\newpage

```{r fig-roc, fig.width=4.5, fig.height=3.0, out.width="60%", fig.cap="Figure 3. ROC curve for the logistic model; the dashed line is random guessing."}
ggplot(roc, aes(x = FPR, y = TPR)) +
  geom_line(color = "#2C7FB8", linewidth = 0.8) +
  geom_abline(linetype = "dashed", color = "grey50") +
  labs(title = paste0("ROC (AUC = ", round(auc, 3), ")"),
       x = "False Positive Rate", y = "True Positive Rate") +
  theme_minimal(base_size = 8)
```

# Section 4 - Limitations and Future Work

**Limitations.** The biggest limit is that we are missing the main causes of delay. Weather is the
clearest example: storms, snow, and fog cause many of the largest delays, but our data has no
weather information at all. The same is true for airport congestion and prior flight delay the
knock on effect. Because these strong factors are missing, our models explain only a small share
of delay, and some of their effect may be hidden inside the features we do have for example,
Time_Block partly stands in for delays that built up earlier in the day. One other limit
matters: the results come from a single year 2015 of US domestic flights, so they may not hold for
other years, other countries, or international routes. The 15 minute line for "delayed" is a useful
standard but still just a cut on a smooth range.

**Future work.** With **one more month**, we would replace our single train/test split with k-fold
cross validation for more stable scores, and add a "previous flight delay" feature to measure the
knock on effect directly. With **three more months**, we would bring in outside weather and
airport congestion data and try tree based models such as random forests to see how much more of
the delay we can explain while keeping the same honest approach to evaluation.
